---
title: "Modeling and Accounting for LD Reference Mismatch in Summary Statistics Fine Mapping"
author: "Gao Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling and Accounting for LD Reference Mismatch in Summary Statistics Fine Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,comment = "#",fig.width = 5,
                      fig.height = 3,fig.align = "center",
                      fig.cap = "&nbsp;",dpi = 120)
```

This vignette gives a practical workflow for using `susie_rss` when the
GWAS summary statistics and the supplied LD matrix may not come from
exactly the same samples. In `susie_rss`, this LD matrix is supplied
through the argument `R`. This vignette complements the
[summary statistics vignette](finemapping_summary_statistics.html) and
the older [RSS diagnostic vignette](susierss_diagnostic.html).

The model introduced here targets a specific mechanism by which LD
mismatch affects SuSiE-RSS: each single effect update subtracts fitted
effects through the supplied LD matrix. If that LD matrix comes from a
finite reference panel, or from a reference panel whose population
differs from the target GWAS population, the subtraction is imperfect.
The leftover residual can then look like additional sparse association
signals.

The approach here is to detect this LD mismatch pattern, quantify the
uncertainty it introduces, and attenuate signals that depend strongly on
uncertain LD subtraction. In the current `susie_rss` implementation,
`R_finite` accounts for finite LD reference panel uncertainty, and
`R_mismatch = "eb"` estimates an additional LD mismatch component from
the data. A separate diagnostic flags residual artifacts that should be
inspected rather than absorbed into the uncertainty adjustment.

```{r}
library(susieR)

count_cs <- function(fit) {
  if (is.null(fit$sets) || is.null(fit$sets$cs)) return(0L)
  length(fit$sets$cs)
}
```

## Recap: a detectable allele coding artifact

The older RSS diagnostic vignette discusses allele coding artifacts in
more detail. We briefly revisit the same idea here because it provides
a useful contrast with the LD mismatch model developed below. The
package contains a small example with one association signal and one
deliberately misaligned variant.

```{r}
data(SummaryConsistency, package = "susieR")
if (!exists("SummaryConsistency"))
  load(file.path("..", "data", "SummaryConsistency.RData"))
attach(SummaryConsistency)

zflip <- z
Rtoy <- ldref
ntoy <- 10000
btoy <- numeric(length(zflip))
btoy[signal_id] <- zflip[signal_id]
```

With the allele coding problem present, ordinary `susie_rss` reports
the true signal and an additional signal at the mismatched variant.

```{r}
fit_toy_bad <- susie_rss(zflip, Rtoy, n = ntoy)
susie_plot(fit_toy_bad, y = "PIP", b = btoy, add_legend = TRUE,
           main = "with allele coding artifact")
points(SummaryConsistency$flip_id,
       fit_toy_bad$pip[flip_id], col = 7, pch = 16)
```

The kriging diagnostic highlights the discordant variant.

```{r}
kr_toy <- kriging_rss(zflip, Rtoy, n = ntoy)
kr_toy$plot
```

After correcting that allele encoding, the extra credible set
disappears.

```{r}
zfix <- zflip
zfix[flip_id] <- -zfix[flip_id]
fit_toy_fixed <- susie_rss(zfix, Rtoy, n = ntoy)
susie_plot(fit_toy_fixed, y = "PIP", b = btoy, add_legend = TRUE,
           main = "after allele correction")
detach(SummaryConsistency)
```

In practice, it is useful to run this basic check before modeling LD
mismatch. When there is an obvious low level allele coding artifact, fix
it directly through **proper bioinformatics procedures**. This can be done using many bioinformatics tools,
for example `pecotmr::allele_qc` from [this package](https://github.com/StatFunGen/pecotmr).
In this toy example, `susie_rss` reports
`r count_cs(fit_toy_bad)` credible sets before the correction and
`r count_cs(fit_toy_fixed)` after the correction, and the kriging rule
identifies the deliberately misaligned variant.
The example from real data below shows a different situation where the
kriging outlier check does not find isolated variants to remove, but
model-based LD uncertainty still changes the fine mapping result.

## Example using real GWAS data and a 1000 Genomes European LD reference panel

The example data are summary statistics from a real gene region.
The supplied LD matrix `R` was computed from a 1000 Genomes European
LD reference panel with roughly 500 samples. The full region used in
the original analysis contains more than 5,000 variants.

In the full region, ordinary `susie_rss` reported 29 credible sets,
whereas setting `R_finite = 500` reduced this to 3 credible sets, and
adding `R_mismatch = "eb"` reduced it to 1 credible set. To keep the
vignette small, the example included here uses a subset of 500 variants.
The rest of this vignette walks through this smaller example step by
step, showing how to use `R_finite`, `R_mismatch`, and the accompanying
diagnostics to evaluate LD mismatch in practice.


```{r}
data(rss_mismatch_example, package = "susieR")
if (!exists("rss_mismatch_example"))
  load(file.path("..", "data", "rss_mismatch_example.RData"))
attach(rss_mismatch_example)
str(rss_mismatch_example)
```

```{r}
length(z)
dim(R)
```

The marginal z-scores show the association pattern before any
fine mapping model is fitted.

```{r}
susie_plot(z, y = "z")
```

The marginal z-scores show one apparent signal spread over many
variants. In this subset, the smallest marginal p-value is approximately
$6.0 \times 10^{-4}$.

We define a few small helpers for the comparisons below.

```{r}
diag_scalar <- function(fit, name) {
  d <- fit$R_finite_diagnostics
  if (is.null(d) || is.null(d[[name]])) return(NA_real_)
  as.numeric(d[[name]][1])
}

diag_flag <- function(fit, name) {
  d <- fit$R_finite_diagnostics
  if (is.null(d) || is.null(d[[name]])) return(NA)
  as.logical(d[[name]][1])
}

fit_table <- function(fits) {
  data.frame(
    method = names(fits),
    n_cs = vapply(fits, count_cs, integer(1)),
    niter = vapply(fits, function(x) x$niter, integer(1)),
    converged = vapply(fits, function(x) isTRUE(x$converged), logical(1)),
    Q_art = vapply(fits, diag_scalar, numeric(1), name = "Q_art"),
    R_sensitivity_flag = vapply(fits, diag_flag, logical(1),
                                name = "R_sensitivity_flag"),
    R_reliability_flag = vapply(fits, diag_flag, logical(1),
                                name = "R_reliability_flag")
  )
}
```

## Ordinary SuSiE-RSS

First run `susie_rss` without any LD mismatch correction.

```{r}
fit_rss <- susie_rss(bhat = bhat, shat = shat,
                     R = R, n = n)
count_cs(fit_rss)
```

```{r}
susie_plot(fit_rss, y = "PIP", add_legend = TRUE,
           main = "susie_rss")
```

```{r}
fit_rss$sets$purity
cs_corr <- get_cs_correlation(fit_rss, Xcorr = R)
round(cs_corr, 3)
```

In this subset, the ordinary fit reports `r count_cs(fit_rss)`
credible sets and does not converge within `r fit_rss$niter`
iterations. The reported credible sets are internally tight, with
minimum purity at least
`r round(min(fit_rss$sets$purity$min.abs.corr), 3)`, but several of
them are nearly perfectly correlated with each other. This combination,
many high-purity credible sets that are mutually correlated, is a
typical sign of LD mismatch.

## Kriging diagnostic

The older RSS diagnostic checks whether individual z-scores are
surprising conditional on the other z-scores and the supplied LD matrix.
This is useful for finding isolated allele coding problems. It is
described in more detail in the
[RSS diagnostic vignette](susierss_diagnostic.html).

```{r}
kr <- kriging_rss(z, R, n = n)
kr$plot
```

The default rule labels possible allele switches when `logLR > 2` and
`abs(z) > 2`.

```{r}
outlier <- which(kr$conditional_dist$logLR > 2 &
                 abs(kr$conditional_dist$z) > 2)
length(outlier)
```

For this example, the kriging diagnostic identifies `r length(outlier)`
isolated allele switch outliers. This illustrates a common distinction:
kriging is good for obvious, localized allele coding artifacts, but this
example is not of that type. It is not designed to model the broader
finite LD reference error and mismatch between the LD reference panel
and the target GWAS population that can affect an entire region.

## Finite LD reference panel correction

The LD matrix here was computed from an LD reference panel of about 500
individuals. Setting `R_finite = 500` tells `susie_rss` to account for
this finite LD reference panel uncertainty in each SER update.

```{r}
fit_B500 <- susie_rss(bhat = bhat, shat = shat,
                      R = R, n = n,
                      R_finite = 500)
count_cs(fit_B500)
```

```{r}
susie_plot(fit_B500, y = "PIP", add_legend = TRUE,
           main = "R_finite = 500")
```

```{r}
vapply(fit_B500$sets$cs, length, integer(1))
fit_B500$sets$purity
cs_corr_B500 <- get_cs_correlation(fit_B500, Xcorr = R)
round(cs_corr_B500, 3)
```

In this subset of 500 variants, the finite LD reference correction
reduces the result from `r count_cs(fit_rss)` credible sets to
`r count_cs(fit_B500)`. In the original larger region, using the same
correction reduced the ordinary `susie_rss` result from 29 credible
sets to 3. The credible sets also become less artificially sharp. Their
sizes change from 2, 9, 9, 11 and 9 in the ordinary fit to 20, 3, 46 and
46, and the lowest purity drops from 0.9831 to 0.9496. The CS
correlations are less uniformly extreme, but some remain high: one pair
is still 1.000 and another is 0.978. Thus `R_finite = 500` helps, but it
does not fully resolve the mismatch pattern.

## Empirical Bayes mismatch correction

Finite LD reference uncertainty only accounts for the fact that the LD
matrix was estimated from a finite panel. It does not account for
systematic mismatch between the LD reference population and the GWAS
population. The option `R_mismatch = "eb"` estimates an additional
variance component shared across the region from the fitted residuals
and uses it in later SER updates.

```{r}
fit_B500_eb <- susie_rss(bhat = bhat, shat = shat,
                         R = R, n = n,
                         R_finite = 500, R_mismatch = "eb")
count_cs(fit_B500_eb)
```

```{r}
susie_plot(fit_B500_eb, y = "PIP", add_legend = TRUE,
           main = "R_finite = 500, R_mismatch = eb")
```

Here, adding `R_mismatch = "eb"` reduces the result to
`r count_cs(fit_B500_eb)` credible set and the fit converges.

One can also run the mismatch model without specifying a finite
LD reference panel size. This treats the continuous mismatch variance as
the only LD uncertainty component.

```{r}
fit_Binf_eb <- susie_rss(bhat = bhat, shat = shat,
                         R = R, n = n,
                         R_mismatch = "eb")
count_cs(fit_Binf_eb)
```

```{r}
susie_plot(fit_Binf_eb, y = "PIP", add_legend = TRUE,
           main = "R_mismatch = eb")
```

This fit also reports `r count_cs(fit_Binf_eb)` credible set and
converges quickly.

```{r}
vapply(fit_B500_eb$sets$cs, length, integer(1))
fit_B500_eb$sets$purity
max(fit_B500_eb$pip)
```

In this subset, the two EB mismatch fits report the same 95% credible
set. It contains 34 variants, has minimum purity 0.8609, and has
coverage 0.9543; the largest PIP is 0.1142. Compared with the earlier
fits, the EB adjustment keeps one signal but represents it as a broad
credible set rather than several small, highly correlated credible sets.
This is more consistent with the weak marginal association in this
region.

## Diagnostic metrics

The warnings above should not be ignored. Even after the EB adjustment,
`susie_rss` reports diagnostic warnings, indicating that the fine
mapping result remains uncertain. The fit object returns several
diagnostic metrics to summarize this uncertainty.

```{r}
fits <- list(susie_rss = fit_rss,
             R_finite_500 = fit_B500,
             R_finite_500_eb = fit_B500_eb,
             R_mismatch_eb = fit_Binf_eb)
knitr::kable(fit_table(fits), digits = 3)
```

There are two relevant reliability diagnostics. `Q_art` measures
whether the final residual points into directions where the supplied LD
matrix has little support. `R_sensitivity_flag` measures whether reported
credible sets depend strongly on the LD uncertainty penalty. If
either diagnostic is triggered, `R_reliability_flag` is `TRUE`, and
the fit should be interpreted cautiously.

In this subset, `R_finite = 500` is flagged by the credible set
sensitivity diagnostic. The two EB mismatch fits are flagged by the
residual space diagnostic, even though they report only one credible
set. This is why the warning points users to
`fit$R_finite_diagnostics` rather than treating the corrected fit as
automatically reliable.

## Diagnostic with one effect

When the reliability flag is raised, the fit object may contain a
diagnostic model with one effect:

```{r}
ser <- fit_Binf_eb$R_finite_diagnostics$ser_model
names(ser)
```

This is a credible set model with one effect as in Maller et al.
(2012). For this model, the single row of `alpha` is the PIP.

```{r}
identical(as.numeric(ser$alpha[1, ]), as.numeric(ser$pip))
count_cs(ser)
vapply(ser$sets$cs, length, integer(1))
ser$sets$purity
max(ser$pip)
identical(ser$sets$cs, fit_Binf_eb$sets$cs)
isTRUE(all.equal(as.numeric(ser$pip), as.numeric(fit_Binf_eb$pip)))
```

```{r}
susie_plot(ser, y = "PIP", add_legend = TRUE,
           main = "diagnostic with one effect")
```

In this example, the one-effect diagnostic gives the same result as the
EB mismatch fit above: one 34-variant credible set, minimum purity
0.8609, and maximum PIP 0.1142. This exact match is a coincidence, not
a general property of the method. When an EB mismatch fit produces a
reliability warning, compare the EB result with this one-effect
diagnostic rather than treating the EB fit as automatically reliable.

## Summary

For routine analyses with GWAS summary statistics and an external LD
reference panel:

- First run basic allele and variant alignment checks. The kriging
  diagnostic is useful for obvious, localized allele coding artifacts.
- Fit ordinary `susie_rss` and look for signs of instability, such as
  nonconvergence, multiple highly correlated credible sets, or many
  small credible sets in a weak association region.
- Use `R_finite` when the LD matrix was estimated from a finite LD
  reference panel.
- Add `R_mismatch = "eb"` when broader LD mismatch remains after
  accounting for finite reference uncertainty.
- Inspect `fit$R_finite_diagnostics`, including `Q_art`,
  `R_sensitivity_flag`, and `R_reliability_flag`.
- If a reliability warning is raised, compare the EB result with the
  one-effect diagnostic when available, and interpret the fine mapping
  result conservatively.

## Caution: our method is not gauranteed to capture allele coding artifacts

The toy example above is based on the out-of-sample LD reference example
from Zou et al. (2022). Because its LD matrix was computed from a finite
reference panel of 500 individuals, it is appropriate to set
`R_finite = 500` when using the new diagnostics. This helps flag the
problem, but it still does not replace allele alignment QC.

```{r}
fit_toy_bad_B500 <- susie_rss(zflip, Rtoy, n = ntoy, R_finite = 500)
fit_toy_bad_B500_eb <- susie_rss(zflip, Rtoy, n = ntoy,
                                 R_finite = 500,
                                 R_mismatch = "eb")
fit_toy_bad_eb <- susie_rss(zflip, Rtoy, n = ntoy,
                            R_mismatch = "eb")

knitr::kable(data.frame(
  method = c("ordinary susie_rss",
             "R_finite = 500",
             "R_finite = 500, R_mismatch = eb",
             "R_mismatch = eb",
             "after allele correction"),
  n_cs = c(count_cs(fit_toy_bad),
           count_cs(fit_toy_bad_B500),
           count_cs(fit_toy_bad_B500_eb),
           count_cs(fit_toy_bad_eb),
           count_cs(fit_toy_fixed)),
  flipped_variant_pip = c(fit_toy_bad$pip[SummaryConsistency$flip_id],
                          fit_toy_bad_B500$pip[SummaryConsistency$flip_id],
                          fit_toy_bad_B500_eb$pip[SummaryConsistency$flip_id],
                          fit_toy_bad_eb$pip[SummaryConsistency$flip_id],
                          fit_toy_fixed$pip[SummaryConsistency$flip_id]),
  R_reliability_flag = c(NA,
                         fit_toy_bad_B500$R_finite_diagnostics$R_reliability_flag,
                         fit_toy_bad_B500_eb$R_finite_diagnostics$R_reliability_flag,
                         fit_toy_bad_eb$R_finite_diagnostics$R_reliability_flag,
                         NA)
), digits = 3)
```

```{r}
susie_plot(fit_toy_bad_B500_eb, y = "PIP", b = btoy, add_legend = TRUE,
           main = "allele artifact with R_finite = 500, R_mismatch = eb")
points(SummaryConsistency$flip_id,
       fit_toy_bad_B500_eb$pip[SummaryConsistency$flip_id],
       col = 7, pch = 16)
```

With `R_finite = 500`, the fit still reports 2 credible sets and the
flipped variant still has PIP close to 1, but `R_reliability_flag` is
`TRUE`. Adding `R_mismatch = "eb"` gives the same result in this
example. In contrast, using `R_mismatch = "eb"` without `R_finite = 500`
does not even flag the artifact. Directly correcting the allele encoding
is what reduces the fit to 1 credible set.

```{r cleanup, include=FALSE}
detach(rss_mismatch_example)
```

## Session information

```{r session-info}
sessionInfo()
```

## Reference

Maller JB, McVean G, Byrnes J, Vukcevic D, Palin K, Su Z, et al.
Bayesian refinement of association signals for 14 loci in 3 common
diseases. Nature Genetics 44, 1294-1301 (2012).
