---
title: "Diagnosing and modeling R-reference mismatch in SuSiE-RSS"
author: "Gao Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnosing and modeling R-reference mismatch in SuSiE-RSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,comment = "#",fig.width = 5,
                      fig.height = 3,fig.align = "center",
                      fig.cap = "&nbsp;",dpi = 120)
```

This vignette gives a practical workflow for using `susie_rss` when the
summary statistics and the supplied correlation matrix `R` may not come
from exactly the same samples. It complements the
[summary-statistics vignette](finemapping_summary_statistics.html) and
the older [RSS diagnostic vignette](susierss_diagnostic.html), and
focuses on the model-based corrections controlled by `R_finite` and
`R_mismatch`.

The basic issue is simple. In SuSiE-RSS, each single-effect update
uses residual summary statistics after subtracting the fitted effects
through `R`. If the supplied `R` is estimated from a finite reference
panel, or from a source population that differs from the target GWAS
population, this subtraction is imperfect. The remaining residual can
then look like additional sparse signals, leading to too many credible
sets.

The new RSS mismatch model treats two parts of this problem as
variance. The finite reference-panel component is specified by
`R_finite`, and the additional source-target mismatch component is
estimated by `R_mismatch = "eb"`. A separate diagnostic checks for
residual artifacts that should not simply be absorbed into the
variance model.

```{r}
library(susieR)

count_cs <- function(fit) {
  if (is.null(fit$sets) || is.null(fit$sets$cs)) return(0L)
  length(fit$sets$cs)
}
```

## Warm-up: an isolated allele-coding artifact

The older RSS diagnostic is most useful when a small number of
variants have clear allele-coding problems. The package contains a
small example with one association signal and one deliberately
misaligned variant.

```{r}
data(SummaryConsistency, package = "susieR")
if (!exists("SummaryConsistency"))
  load(file.path("..", "data", "SummaryConsistency.RData"))
attach(SummaryConsistency)

zflip <- z
Rtoy <- ldref
ntoy <- 10000
btoy <- numeric(length(zflip))
btoy[signal_id] <- zflip[signal_id]
```

With the allele-coding problem present, ordinary `susie_rss` reports
the true signal and an additional signal at the mismatched variant.

```{r}
fit_toy_bad <- susie_rss(zflip, Rtoy, n = ntoy)
susie_plot(fit_toy_bad, y = "PIP", b = btoy, add_legend = TRUE,
           main = "with allele-coding artifact")
points(SummaryConsistency$flip_id,
       fit_toy_bad$pip[flip_id], col = 7, pch = 16)
```

The kriging diagnostic highlights the discordant variant.

```{r}
kr_toy <- kriging_rss(zflip, Rtoy, n = ntoy)
kr_toy$plot
```

After correcting that allele encoding, the extra credible set
disappears.

```{r}
zfix <- zflip
zfix[flip_id] <- -zfix[flip_id]
fit_toy_fixed <- susie_rss(zfix, Rtoy, n = ntoy)
susie_plot(fit_toy_fixed, y = "PIP", b = btoy, add_legend = TRUE,
           main = "after allele correction")
detach(SummaryConsistency)
```

This warm-up is an important baseline: when there is an obvious local
allele-coding artifact, fix it directly. In this toy example,
`susie_rss` reports `r count_cs(fit_toy_bad)` credible sets before
the correction and `r count_cs(fit_toy_fixed)` after the correction,
and the kriging rule identifies the deliberately misaligned variant.
The real-data example below shows a different situation where the
kriging outlier check does not find isolated variants to remove, but
model-based reference uncertainty still changes the fine-mapping
result.

## A real-data example

The example data are summary statistics from one real-data gene region.
The supplied correlation matrix `R` was computed from a real-world 1000
Genomes European reference panel with roughly 500 samples. To keep the
vignette small, we use a 500-variant subset from a larger region. In
the full region, ordinary `susie_rss` reported 29 credible sets,
whereas setting `R_finite = 500` reduced this to 3 credible sets, and
adding `R_mismatch = "eb"` reduced it to 1 credible set.

The data object included here is intentionally minimal and name-free:
it contains only z-scores, regression coefficients, standard errors,
the reference sample-size estimate, and the reference correlation
matrix.

```{r}
data(rss_mismatch_example, package = "susieR")
if (!exists("rss_mismatch_example"))
  load(file.path("..", "data", "rss_mismatch_example.RData"))
attach(rss_mismatch_example)
str(rss_mismatch_example)
```

```{r}
length(z)
dim(R)
```

The marginal z-scores show the association pattern before any
fine-mapping model is fitted.

```{r}
susie_plot(z, y = "z")
```

We define a few small helpers for the comparisons below.

```{r}
diag_scalar <- function(fit, name) {
  d <- fit$R_finite_diagnostics
  if (is.null(d) || is.null(d[[name]])) return(NA_real_)
  as.numeric(d[[name]][1])
}

diag_flag <- function(fit, name) {
  d <- fit$R_finite_diagnostics
  if (is.null(d) || is.null(d[[name]])) return(NA)
  as.logical(d[[name]][1])
}

fit_table <- function(fits) {
  data.frame(
    method = names(fits),
    n_cs = vapply(fits, count_cs, integer(1)),
    niter = vapply(fits, function(x) x$niter, integer(1)),
    converged = vapply(fits, function(x) isTRUE(x$converged), logical(1)),
    Q_art = vapply(fits, diag_scalar, numeric(1), name = "Q_art"),
    R_sensitivity_flag = vapply(fits, diag_flag, logical(1),
                                name = "R_sensitivity_flag"),
    R_reliability_flag = vapply(fits, diag_flag, logical(1),
                                name = "R_reliability_flag")
  )
}
```

## Ordinary SuSiE-RSS

First run `susie_rss` without any reference-mismatch correction.

```{r}
fit_rss <- susie_rss(bhat = bhat, shat = shat,
                     R = R, n = n,
                     L = 40, L_greedy = 10, max_iter = 50)
count_cs(fit_rss)
```

```{r}
susie_plot(fit_rss, y = "PIP", add_legend = TRUE,
           main = "susie_rss")
```

In this subset, the ordinary fit reports `r count_cs(fit_rss)`
credible sets and does not converge within 50 iterations. This is a
warning sign in a region where the reference panel may not match the
target summary statistics.

## Kriging diagnostic and outlier removal

The older RSS diagnostic checks whether individual z-scores are
surprising conditional on the other z-scores and the supplied `R`.
This is useful for finding isolated allele-coding problems. It is
described in more detail in the
[RSS diagnostic vignette](susierss_diagnostic.html).

```{r}
kr <- kriging_rss(z, R, n = n)
kr$plot
```

The built-in rule labels possible allele switches when `logLR > 2` and
`abs(z) > 2`.

```{r}
outlier <- which(kr$conditional_dist$logLR > 2 &
                 abs(kr$conditional_dist$z) > 2)
length(outlier)
```

For this example, the kriging diagnostic identifies `r length(outlier)`
isolated allele-switch outliers. Removing those outliers therefore does
not change the fit.

```{r}
keep <- setdiff(seq_along(z), outlier)
fit_kriging_removed <- susie_rss(bhat = bhat[keep],
                                 shat = shat[keep],
                                 R = R[keep, keep],
                                 n = n,
                                 L = 40, L_greedy = 10,
                                 max_iter = 50)
count_cs(fit_kriging_removed)
```

The refitted model again reports `r count_cs(fit_kriging_removed)`
credible sets. This illustrates a common distinction. Kriging is good
for obvious, localized allele-coding artifacts. It is not designed to
model the broader finite-reference and source-target mismatch that can
affect an entire region.

## Finite reference-panel correction

The reference correlation matrix here was computed from about 500
individuals. Setting `R_finite = 500` tells `susie_rss` to account for
this finite reference-panel uncertainty in each SER update.

```{r}
fit_B500 <- susie_rss(bhat = bhat, shat = shat,
                      R = R, n = n,
                      L = 40, L_greedy = 10, max_iter = 50,
                      R_finite = 500)
count_cs(fit_B500)
```

```{r}
susie_plot(fit_B500, y = "PIP", add_legend = TRUE,
           main = "R_finite = 500")
```

In this 500-variant subset, the finite-reference correction reduces
the result from `r count_cs(fit_rss)` credible sets to
`r count_cs(fit_B500)`. In the original larger region, using the same
correction reduced the ordinary `susie_rss` result from 29 credible
sets to 3.

## Empirical-Bayes mismatch correction

Finite-reference uncertainty only accounts for the fact that `R` was
estimated from a finite panel. It does not account for systematic
source-target mismatch between the reference population and the GWAS
population. The option `R_mismatch = "eb"` estimates an additional
region-level variance component from the fitted residuals and uses it
in later SER updates.

```{r}
fit_B500_eb <- susie_rss(bhat = bhat, shat = shat,
                         R = R, n = n,
                         L = 40, L_greedy = 10, max_iter = 50,
                         R_finite = 500, R_mismatch = "eb")
count_cs(fit_B500_eb)
```

```{r}
susie_plot(fit_B500_eb, y = "PIP", add_legend = TRUE,
           main = "R_finite = 500, R_mismatch = eb")
```

Here, adding `R_mismatch = "eb"` reduces the result to
`r count_cs(fit_B500_eb)` credible set and the fit converges.

One can also run the mismatch model without specifying a finite
reference-panel size. This treats the continuous mismatch variance as
the only reference-uncertainty component.

```{r}
fit_Binf_eb <- susie_rss(bhat = bhat, shat = shat,
                         R = R, n = n,
                         L = 40, L_greedy = 10, max_iter = 50,
                         R_mismatch = "eb")
count_cs(fit_Binf_eb)
```

```{r}
susie_plot(fit_Binf_eb, y = "PIP", add_legend = TRUE,
           main = "R_mismatch = eb")
```

This fit also reports `r count_cs(fit_Binf_eb)` credible set and
converges quickly.

## What to check

The most useful output for users is not the exact value of the fitted
variance component. Instead, check whether the fit converged, how many
credible sets were reported, and whether the mismatch diagnostics
raised a reliability flag.

```{r}
fits <- list(susie_rss = fit_rss,
             kriging_removed = fit_kriging_removed,
             R_finite_500 = fit_B500,
             R_finite_500_eb = fit_B500_eb,
             R_mismatch_eb = fit_Binf_eb)
fit_table(fits)
```

There are two relevant reliability diagnostics. `Q_art` measures
whether the final residual points into directions where the supplied
`R` has little support. `R_sensitivity_flag` measures whether reported
credible sets depend strongly on the reference-uncertainty penalty. If
either diagnostic is triggered, `R_reliability_flag` is `TRUE`, and
the fit should be interpreted cautiously.

In this subset, `R_finite = 500` is flagged by the credible-set
sensitivity diagnostic. The two EB mismatch fits are flagged by the
residual-space diagnostic, even though they report only one credible
set. This is why the warning points users to
`fit$R_finite_diagnostics` rather than treating the corrected fit as
automatically reliable.

## One-effect diagnostic

When the reliability flag is raised, the fit object may contain a
one-effect diagnostic model:

```{r}
ser <- fit_Binf_eb$R_finite_diagnostics$ser_model
names(ser)
```

This is a one-effect credible-set model as in Maller et al. (2012).
For this one-effect model, the single row of `alpha` is the PIP.

```{r}
identical(as.numeric(ser$alpha[1, ]), as.numeric(ser$pip))
count_cs(ser)
```

```{r}
susie_plot(ser, y = "PIP", add_legend = TRUE,
           main = "one-effect diagnostic")
```

The one-effect diagnostic is not a replacement for careful review of
the summary statistics and reference panel. It is a conservative
reference point when the multi-effect fit is flagged as unreliable.

## Compact workflow

For routine analyses with an out-of-sample reference panel, the core
workflow is:

```{r}
fit <- susie_rss(bhat = bhat, shat = shat,
                 R = R, n = n,
                 L = 40, L_greedy = 10,
                 R_finite = 500,
                 R_mismatch = "eb")

count_cs(fit)
fit$converged
fit$R_finite_diagnostics$R_reliability_flag
```

If a reliability warning is raised, inspect
`fit$R_finite_diagnostics`, compare with the one-effect diagnostic
when available, and consider improving the reference panel or using a
larger better-matched reference panel. A larger reference panel reduces
the finite-reference penalty. A better-matched reference panel reduces
the population-mismatch component that `R_mismatch = "eb"` has to
absorb.

The older kriging diagnostic remains useful for localized allele
problems, and is demonstrated in the
[RSS diagnostic vignette](susierss_diagnostic.html). The model-based
correction here addresses a different failure mode: residual signal
created because the fitted effects are subtracted using an imperfect
reference correlation matrix.

```{r cleanup, include=FALSE}
detach(rss_mismatch_example)
```

## Session information

```{r session-info}
sessionInfo()
```

## Reference

Maller JB, McVean G, Byrnes J, Vukcevic D, Palin K, Su Z, et al.
Bayesian refinement of association signals for 14 loci in 3 common
diseases. Nature Genetics 44, 1294-1301 (2012).
